Rao-Blackwellized smoother for Conditionally Linear and Gaussian models

### 2016-11-15 TexArxiv/clgm.tex
MAJ

The paper is organized as follows. The proposed modification of the fully Rao Blackwellised Monte Carlo smoothers are presented in Section~\ref{sec:RaoBlackwell}. The performance of all these methods is illustrated in Section~\ref{sec:numerical:experiments} with simulated data. In Section~\ref{}, an application of regime switchings for commodity markets is presented using CME crude oil (WTI) data. A detailed derivation of the algorithms is provided in the Appendix.

@@

\section{Numerical experiments}
\label{sec:numerical:experiments}
\subsection{Simulated data}
This section highlights the performance of the additional Rao Blackwellisation steps for the two-filter and the FFBS approximations of the marginal smoothing distributions in the case where $J=2$. In each setting the transition matrix $Q$ is such that the states are prone to stay in the same regime  for many time steps as expected for the WTI crude oil data,  see Section~\ref{sec:exp}. First, the FFBS algorithm is applied to a simple one-dimensional model with:
\[
\pi(1)=\pi(2) = 1/2\\;\;d_1 = 0.5\;\; d_2 = 0;\;c_1= c_2 = 0 \eqsp,
\]
\[
Q = \begin{pmatrix}0.99 & 0.01\\ 0.03 & 0.97\end{pmatrix}\;\;T_1 = T_2 = 1\;\;\barH_1 = \barH_2= 0.05\eqsp,
\]
\[
B_1 = B_2 =1\;\;\barG_1 = 0.3\;\;\barG_2 = 0.1\eqsp.
\]
The two FFBS algorithms are used with $N = \tilde{N} = 50$ and the estimated posterior probabilities $\mathbb{P}(a_k=1|Y_{1:n})$ are compared to the estimation obtained with the proposed FFBS algorithm with rejuvenation and $2000$ particles. Figure~\ref{fig:ffbs:posterior} illustrates that the algorithm with rejuvenation slightly improves the original algorithm in this simple case. 
\begin{figure}
\centering
\includegraphics[scale=.3]{FFBSplot.eps}
\caption{Estimated posterior probabilities by the two FFBS algorithms.}
\label{fig:ffbs:posterior}
\end{figure}
Figure~\ref{fig:ffbs:err} displays the mean estimation error over $100$ independent Monte Carlo runs. The estimation error is defined as the difference between the the estimation obtained with the proposed FFBS algorithm with rejuvenation and $2000$ particles and the estimation given by the two FFBS algorithms with $N = \tilde{N} = 50$.
\begin{figure}
\centering
\includegraphics[scale=.3]{err_est_ffbs.eps}
\caption{Posterior probabilities estimation error fo the two FFBS algorithms.}
\label{fig:ffbs:err}
\end{figure}
%In addition, Table~\ref{fig:ffbs:RMSE} displays the RMSE for the estimation of the labels as a function of time with $100$ independent batches of data and shows that the additional rejuvenation step improves the performance of the algorithm (at time $i$, the estimated label is set to $1$ if the estimation of $\mathbb{P}(a_i|Y_{1:n})$ is greater than $0.5$ and to $2$ otherwise).

@@

\begin{figure}
\centering
\includegraphics[scale=.5]{N15_1_500}
\includegraphics[scale=.5]{N15_501_1000}
\includegraphics[scale=.5]{N15_1001_1500}
\caption{Experiment 2 - estimated probabilites by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=15$ particles, RBTF with $N=1000$ particles (black line) and true labels (squared black).}
\label{fig:exp2:label}
\end{figure}

\subsection{Application to CME crude oil (WTI)}
\label{sec:exp}

\subsubsection*{State dynamics}

@@

Statistical estimation is performed using market observations and cannot be done directly under the risk-neutral probability but under the historical probability $\mathbb{P}$. It is assumed that each factor (spot, convenience yield) has its own constant market price of risk, $\lambda^S$ for $S_t$ and $\lambda^{\delta}$ for $\delta_t$. The historical dynamics of the Gibson-Schwartz model is then described as follows:
\rmd S_{t} & = (\mu -\delta_{t})S_{t}\rmd t+\sigma_{a_t} S_{t} \rmd \widetilde{W}_{t}^{1} \eqsp,\\
\rmd \delta_{t} & = \kappa_{a_t}(\widetilde{\alpha}_{a_t}-\delta_{t})\rmd t+\eta_{a_t} \rmd \widetilde{W}_{t}^{2} \eqsp, \quad \rmd \langle \widetilde{W}_t^1, \widetilde{W}_t^2\rangle = \rho \rmd t\eqsp,
where $( (\widetilde{W}_t^1,\widetilde{W}_t^2), t \geq 0)$ are standard Brownian motions under the historical probability $\mathbb{P}$ with  correlation $\rho_{a_t}$, $\mu = r - \lambda^S$ and $\widetilde{\alpha}_{a_t} = \alpha_{a_t} - \lambda^{\delta}/\kappa_{a_t} $. This model allows to exhibit fundamental features of commodity future prices, which typically display different regimes of volatility and/or convenience yield. A two-regime model is already sufficient to produce stylized effects such as contango (increase of future prices) and backwardation (decrease of future prices).  Assuming that the switching rate between regimes is small compared to the inverse of the discretization period, the discretized version of the spot price and convenience yield $Z_i= (\ln S_i, \delta_i)$ (the sampling period is taken to be 1) is modeled as a Conditionally Linear Gaussian Model. The explicit integration of this SDE detailed in Lemma~\ref{lem:integratedSDE} yields the following discrete time model for $(Z_i)_{i\ge 0}$:

@@

\begin{pmatrix} \left[\mu- \widetilde{\alpha}_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \Delta t_i + \widetilde{\alpha}_{a_{i-1}}[1-e^{-\kappa_{a_{i-1}} \Delta t_i}]/\kappa_{a_{i-1}}\\
\widetilde{\alpha}_{a_{i-1}} [1-e^{-\kappa_{a_{i-1}} \Delta t_i}] \end{pmatrix}\eqsp,\\

@@

\begin{pmatrix} 1 & -[1-e^{-\kappa_{a_{i-1}} \Delta t_i}]/\kappa_{a_{i-1}} \\ 0 & e^{-\kappa_{a_{i-1}} \Delta t_i} \end{pmatrix} \eqsp,\\
\barH_{a_{i-1}}(1,1) &= \sigma^2_{a_{i-1}} \Delta t_i + \eta^2_{a_{i-1}}\left\{\Delta t_i + (1-e^{-2\kappa_{a_{i-1}} \Delta t_i})/(2\kappa_{a_{i-1}}) - 2(1-e^{-\kappa \Delta t_i})/\kappa_{a_{i-1}}\right\}/\kappa_{a_{i-1}}^2 \\
&\hspace{5.2cm} - 2\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}\left\{ \Delta t_i - (1-e^{-\kappa_{a_{i-1}} \Delta t_i})/\kappa_{a_{i-1}} \right\}/\kappa\eqsp, \\
\barH_{a_{i-1}}(1,2) & = \left(\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}-\eta^2_{a_{i-1}}/\kappa_{a_{i-1}}\right)\left(1-e^{-\kappa_{a_{i-1}}\Delta t_i} \right)/\kappa_{a_{i-1}} \\
&\hspace{5.2cm}+ \eta^2_{a_{i-1}}\left(1-e^{-2\kappa_{a_{i-1}} \Delta t_i} \right)/(2\kappa_{a_{i-1}}^2)\eqsp,\\

@@

\barH_{a_{i-1}}(2,2) &= \eta^2_{a_{i-1}}\left( 1-e^{-2\kappa_{a_{i-1}}\Delta t_i} \right)/(2\kappa_{a_{i-1}})\eqsp.

@@

\subsubsection*{Observation model}

@@

\subsubsection*{Numerical results}
\label{sec:crudeoil}
%Following \cite{carmona:ludkovski:2004}, under the risk neutral probability $\mathbb{Q}$, we assume that the spot price $S_t$ of the Chicago Mercantile Exchange (CME) crude oil and its associated convenience yield $\delta_t$ are solutions to the SDE:
%\begin{align*}
%\rmd S_t &= \left(\tilde{\mu}_{a_{t}} - \delta_t\right) S_t \rmd t + \sigma_{a_t} S_t \rmd \tilde{W}^1_t \eqsp,\\
%\rmd \delta_t &= \kappa \left( \tilde{\alpha}_{a_t} - \delta_t \right) \rmd t + \eta_{a_t} \rmd \tilde{W}^2_t \eqsp,
%\end{align*}
%where $\tilde{W}^1$ and $\tilde{W}^2$ are two correlated Brownian motions under the historical probability $\mathbb{P}$.
%
%In this case,  $\delta_t$ is an Orstein-Uhlenbeck process.
%
%Under the risk-neutral probability $\mathbb{Q}$, this dynamic becomes:
%\begin{align*}
%\rmd S_t &= \left(r - \delta_t\right)S_t \rmd t + \sigma_{a_t} S_t \rmd W^1_t \eqsp,\\
%\rmd \delta_t &= \left[\kappa \left(\alpha_{a_t} - \delta_t \right) - \lambda_{a_t} \right] \rmd t + \eta_{a_t} \rmd  W^2_t \eqsp,
%\end{align*}
%Define $\alpha_{a_t} \eqdef \tilde{\alpha}_{a_t} - \lambda_{a_t}/\kappa$. Therefore, the dynamic of $Z_t = (\ln(S_t), \delta_t)$ under the risk-neutral mesure is:
%\begin{align*}
%\rmd \ln(S_t) &= \left(r - \delta_t- \sigma^2_{a_t}/2 \right) dt + \sigma_{a_t} \rmd W^1_t \eqsp,\\
%\rmd \delta_t &= \kappa\left( \alpha_{a_t} - \delta_t\right)dt + \eta_{a_t} \rmd W^2_t \eqsp,
%\end{align*}
%where $(a_t)_{t\ge 0}$ is a Markov process taking values in  $\{1,\ldots,J\}$ indicating the market regime and $W^1$ and $W^2$ are Brownian motions under $\mathbb{Q}$ such that $\rmd<W^1, W^2>_t = \rho_{a_t}\rmd t$. The interest rate $r$ is constant and $\lambda_{a_t}$ is a risk associated with the convenience yield dynamic when the market regime is $a_t$. In the following, it is assumed that between two consecutive observations at times $t_{i-1}$ and $t_{i}$, the market regime is fixed so that the solutions to the previous SDE may be computed explicitly if $a_t$ is set to $a_{i-1} = a_{t_{i-1}}$. Therefore, it is assumed that the regime indicators are given at times $(t_i)_{i\ge 1}$ by a discrete time Markov chain $(a_{i})_{i\ge 1}$ and do not change between $t_{i-1}$ and $t_{i}$.
%The explicit integration of \eqref{eq:modele} between $t_{i-1}$ and $t_i$ under the approximation that there is no regime change between these two successive time instants provides the following values of the parameter of the discrete time model:
%for $Z_i \eqdef (\ln (S_{t_i}),\delta_{t_i})$:
%\[
%Z_i = d_{a_{i-1}} + T Z_{i-1} + H_{a_{i-1}} \varepsilon_i\eqsp,
%\]
%with $\varepsilon_i$ a standard $2$-dimensional Gaussian random variable and parameters defined as

>>>>

The paper is organized as follows. The proposed modification of the fully Rao Blackwellised Monte Carlo smoothers are presented in Section~\ref{sec:RaoBlackwell}. The performance of all these methods is illustrated in Section~\ref{sec:numerical:experiments} with simulated data. In Section~\ref{sec:crudeoil}, an application of regime switchings for commodity markets is presented using CME crude oil (WTI) data. A detailed derivation of the algorithms is provided in the Appendix.

@@

\section{Simulated data}
\label{sec:numerical:experiments}
This section highlights the performance improvement brought by  the additional Rao-Blackwellisation steps for the two-filter and the FFBS approximations of the marginal smoothing distributions in the case where the number of states is $J=2$. In each setting the transition matrix $Q$ is such that the probability of switching from one regime to the other is small, as expected for the WTI crude oil data;  see Section~\ref{sec:exp}. First, the FFBS algorithm is applied to a simple one-dimensional model with:
\[
\pi(1)=\pi(2) = 1/2\\;\;d_1 = 0.5\;\; d_2 = 0;\;c_1= c_2 = 0 \eqsp,
\]
\[
Q = \begin{pmatrix}0.99 & 0.01\\ 0.03 & 0.97\end{pmatrix}\;\;T_1 = T_2 = 1\;\;\barH_1 = \barH_2= 0.05\eqsp,
\]
\[
B_1 = B_2 =1\;\;\barG_1 = 0.3\;\;\barG_2 = 0.1\eqsp.
\]
The two FFBS algorithms are used with $N = \tilde{N} = 50$ and the estimated posterior probabilities $\mathbb{P}(a_k=1|Y_{1:n})$ are compared to the estimation obtained with the proposed FFBS algorithm with rejuvenation and $2000$ particles. Figure~\ref{fig:ffbs:posterior} illustrates that the algorithm with rejuvenation slightly improves the original algorithm in this simple case.
\begin{figure}
\centering
\includegraphics[scale=.3]{FFBSplot}
\caption{Estimated posterior probabilities by the two FFBS algorithms.}
\label{fig:ffbs:posterior}
\end{figure}
Figure~\ref{fig:ffbs:err} displays the mean estimation error over $100$ independent Monte Carlo runs. The estimation error is defined as the difference between the the estimation obtained with the proposed FFBS algorithm with rejuvenation and $2000$ particles and the estimation given by the two FFBS algorithms with $N = \tilde{N} = 50$.
\begin{figure}
\centering
\includegraphics[scale=.3]{err_est_ffbs}
\caption{Posterior probabilities estimation error fo the two FFBS algorithms.}
\label{fig:ffbs:err}
\end{figure}
%In addition, Table~\ref{fig:ffbs:RMSE} displays the RMSE for the estimation of the labels as a function of time with $100$ independent batches of data and shows that the additional rejuvenation step improves the performance of the algorithm (at time $i$, the estimated label is set to $1$ if the estimation of $\mathbb{P}(a_i|Y_{1:n})$ is greater than $0.5$ and to $2$ otherwise).

@@

\begin{figure}
\centering
\includegraphics[scale=.5]{N15_1_500}
\includegraphics[scale=.5]{N15_501_1000}
\includegraphics[scale=.5]{N15_1001_1500}
\caption{Experiment 2 - estimated smoothing probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=15$ particles, RBTF with $N=1000$ particles (black line) and true labels (squared black).}
\label{fig:exp2:label}
\end{figure}

\section{Application to CME crude oil (WTI)}
\label{sec:crudeoil}
\label{sec:exp}

\subsection{Model}

@@

%Statistical estimation is performed using market observations and cannot be done directly under the risk-neutral probability but under the historical probability $\mathbb{P}$. It is assumed that each factor (spot, convenience yield) has its own constant market price of risk, $\lambda^S$ for $S_t$ and $\lambda^{\delta}$ for $\delta_t$. The historical dynamics of the Gibson-Schwartz model is then described as follows:
%\rmd S_{t} & = (\mu -\delta_{t})S_{t}\rmd t+\sigma_{a_t} S_{t} \rmd \widetilde{W}_{t}^{1} \eqsp,\\
%\rmd \delta_{t} & = \kappa_{a_t}(\widetilde{\alpha}_{a_t}-\delta_{t})\rmd t+\eta_{a_t} \rmd \widetilde{W}_{t}^{2} \eqsp, \quad \rmd \langle \widetilde{W}_t^1, \widetilde{W}_t^2\rangle = \rho \rmd t\eqsp,
%where $( (\widetilde{W}_t^1,\widetilde{W}_t^2), t \geq 0)$ are standard Brownian motions under the historical probability $\mathbb{P}$ with  correlation $\rho_{a_t}$, $\mu = r - \lambda^S$ and $\widetilde{\alpha}_{a_t} = \alpha_{a_t} - \lambda^{\delta}/\kappa_{a_t} $. 
This model allows to exhibit fundamental features of commodity future prices, which typically display different regimes of volatility and/or convenience yield. A two-regime model is already sufficient to produce stylized effects such as contango (increase of future prices) and backwardation (decrease of future prices).  Assuming that the switching rate between regimes is negligible compared to the inverse of the discretization period, the discretized version of the spot price and convenience yield $Z_i= (\ln S_i, \delta_i)$ (the sampling period is taken to be 1) is modeled as a Conditionally Linear Gaussian Model. The explicit integration of this SDE detailed in Lemma~\ref{lem:integratedSDE} yields the following discrete time model for $(Z_i)_{i\ge 0}$:

@@

\begin{pmatrix} \left[\mu- \alpha_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \Delta t_i + \alpha_{a_{i-1}}[1-\rme^{-\kappa_{a_{i-1}} \Delta t_i}]/\kappa_{a_{i-1}}\\
\alpha_{a_{i-1}} [1-\rme^{-\kappa_{a_{i-1}} \Delta t_i}] \end{pmatrix}\eqsp,\\

@@

\begin{pmatrix} 1 & -[1-\rme^{-\kappa_{a_{i-1}} \Delta t_i}]/\kappa_{a_{i-1}} \\ 0 & \rme^{-\kappa_{a_{i-1}} \Delta t_i} \end{pmatrix} \eqsp,\\
\barH_{a_{i-1}}(1,1) &= \sigma^2_{a_{i-1}} \Delta t_i + \eta^2_{a_{i-1}}\left\{\Delta t_i + (1-\rme^{-2\kappa_{a_{i-1}} \Delta t_i})/(2\kappa_{a_{i-1}}) - 2(1-\rme^{-\kappa \Delta t_i})/\kappa_{a_{i-1}}\right\}/\kappa_{a_{i-1}}^2 \\
&\hspace{5.2cm} - 2\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}\left\{ \Delta t_i - (1-\rme^{-\kappa_{a_{i-1}} \Delta t_i})/\kappa_{a_{i-1}} \right\}/\kappa\eqsp, \\
\barH_{a_{i-1}}(1,2) & = \left(\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}-\eta^2_{a_{i-1}}/\kappa_{a_{i-1}}\right)\left(1-\rme^{-\kappa_{a_{i-1}}\Delta t_i} \right)/\kappa_{a_{i-1}} \\
&\hspace{5.2cm}+ \eta^2_{a_{i-1}}\left(1-\rme^{-2\kappa_{a_{i-1}} \Delta t_i} \right)/(2\kappa_{a_{i-1}}^2)\eqsp,\\

@@

\barH_{a_{i-1}}(2,2) &= \eta^2_{a_{i-1}}\left( 1-\rme^{-2\kappa_{a_{i-1}}\Delta t_i} \right)/(2\kappa_{a_{i-1}})\eqsp.

@@

@@

\subsection{Numerical results}
%Following \cite{carmona:ludkovski:2004}, under the risk neutral probability $\mathbb{Q}$, we assume that the spot price $S_t$ of the Chicago Mercantile Exchange (CME) crude oil and its associated convenience yield $\delta_t$ are solutions to the SDE:
%\begin{align*}
%\rmd S_t &= \left(\tilde{\mu}_{a_{t}} - \delta_t\right) S_t \rmd t + \sigma_{a_t} S_t \rmd \tilde{W}^1_t \eqsp,\\
%\rmd \delta_t &= \kappa \left( \tilde{\alpha}_{a_t} - \delta_t \right) \rmd t + \eta_{a_t} \rmd \tilde{W}^2_t \eqsp,
%\end{align*}
%where $\tilde{W}^1$ and $\tilde{W}^2$ are two correlated Brownian motions under the historical probability $\mathbb{P}$.
%
%In this case,  $\delta_t$ is an Orstein-Uhlenbeck process.
%
%Under the risk-neutral probability $\mathbb{Q}$, this dynamic becomes:
%\begin{align*}
%\rmd S_t &= \left(r - \delta_t\right)S_t \rmd t + \sigma_{a_t} S_t \rmd W^1_t \eqsp,\\
%\rmd \delta_t &= \left[\kappa \left(\alpha_{a_t} - \delta_t \right) - \lambda_{a_t} \right] \rmd t + \eta_{a_t} \rmd  W^2_t \eqsp,
%\end{align*}
%Define $\alpha_{a_t} \eqdef \tilde{\alpha}_{a_t} - \lambda_{a_t}/\kappa$. Therefore, the dynamic of $Z_t = (\ln(S_t), \delta_t)$ under the risk-neutral mesure is:
%\begin{align*}
%\rmd \ln(S_t) &= \left(r - \delta_t- \sigma^2_{a_t}/2 \right) dt + \sigma_{a_t} \rmd W^1_t \eqsp,\\
%\rmd \delta_t &= \kappa\left( \alpha_{a_t} - \delta_t\right)dt + \eta_{a_t} \rmd W^2_t \eqsp,
%\end{align*}
%where $(a_t)_{t\ge 0}$ is a Markov process taking values in  $\{1,\ldots,J\}$ indicating the market regime and $W^1$ and $W^2$ are Brownian motions under $\mathbb{Q}$ such that $\rmd<W^1, W^2>_t = \rho_{a_t}\rmd t$. The interest rate $r$ is constant and $\lambda_{a_t}$ is a risk associated with the convenience yield dynamic when the market regime is $a_t$. In the following, it is assumed that between two consecutive observations at times $t_{i-1}$ and $t_{i}$, the market regime is fixed so that the solutions to the previous SDE may be computed explicitly if $a_t$ is set to $a_{i-1} = a_{t_{i-1}}$. Therefore, it is assumed that the regime indicators are given at times $(t_i)_{i\ge 1}$ by a discrete time Markov chain $(a_{i})_{i\ge 1}$ and do not change between $t_{i-1}$ and $t_{i}$.
%The explicit integration of \eqref{eq:modele} between $t_{i-1}$ and $t_i$ under the approximation that there is no regime change between these two successive time instants provides the following values of the parameter of the discrete time model:
%for $Z_i \eqdef (\ln (S_{t_i}),\delta_{t_i})$:
%\[
%Z_i = d_{a_{i-1}} + T Z_{i-1} + H_{a_{i-1}} \varepsilon_i\eqsp,
%\]
%with $\varepsilon_i$ a standard $2$-dimensional Gaussian random variable and parameters defined as
